---
title: "NOAA Storm Data Analysis"
author: "Mugenyi Raymond"
date: "7/15/2021"
output: html_document
---
## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, crop and property damage, and preventing such outcomes to the extent possible is a key concern.By using U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database,  two main questions are addressed.These are; "which types of events are most harmful with respect to population health?" and "which types of events have the greatest economic consequences?". In the  analysis, it was found out that TORNADO is the most harmful event type with respect to population health and Flood  has the greatest economic consequences across United States!



## **Data Processing**
The data was downloaded from the url: http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
and it was then unzipped using the bunzip2 function from the R.utils package.

```{r results='hide'}
library(R.utils)
fileurl <-  "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileurl, destfile =  "NOOAstormdata.csv.bz2")
bunzip2("NOOAstormdata.csv.bz2") # unzipping the stormdata
data <-  read.csv("NOOAstormdata.csv") #reading in the data
```

## *Question 1: Across the United States, which types of events are most harmful with respect to population health?*
Here we shall explore and find out which event causes most injuries and fatalities
At first lets select the key variables relevant in this analysis. These are;
EVTYPE , FATALITIES , INJURIES, PROPDMG, CROPDMG, PROPDMGEXP, and CROPDMGEXP.

```{r results='hide'}
library(dplyr)
data <- select(data, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG", "PROPDMGEXP", "CROPDMGEXP" ))
injuries <- aggregate(data$INJURIES, by = list(data$EVTYPE), FUN = sum) %>% 
  rename(event_type = "Group.1", total_injuries = "x") %>% arrange(desc(total_injuries))
```

```{r results='hide'}
#plot
library(ggplot2)
bar_plot_injuries <- ggplot(injuries[1:5,], aes(event_type, total_injuries)) +
  geom_bar(stat = "identity" , aes(fill = event_type)) +
  labs(x = "Type of Event", y = "Total injuries",
       title = "Total injuries by Event Type", 
       caption = "Figure 1: Only the first 5 devastating events are shown.
       Tornado shows highest number of injuries followed by TSTM Wind, 
       Flood,Excessive heat and lastly Lightning.") + 
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(size = 8.5,hjust = 0.5))

```



For the event type causing most fatalities;
```{r results='hide'}
library(dplyr)
fatalities <- aggregate(data$FATALITIES, by = list(data$EVTYPE), FUN = sum) %>% 
  rename(event_type = "Group.1", total_fatalities = "x") %>% arrange(desc(total_fatalities))

# plot
library(ggplot2)
bar_plot_fatalities <- ggplot(fatalities[1:5,], aes(event_type, total_fatalities)) +
  geom_bar(stat = "identity" , aes(fill = event_type)) + 
  labs(x = "Type of Event",
       y = "Total fatalities", 
       title = "Total fatalities by Event Type",
       caption = "Figure 2:Only the first 5 devastating event types are shown.
       Tornado shows the highest number of fatalities followed by Excessive heat,
       Flash flood, Heat, and lastly Lightning.") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(size = 8.5, hjust = 0.5))

```


## *Question 2: Across the United States, which types of events have the greatest economic consequences?*
To address this question, we have to explore the property damange, crop damage and the total damage (property damage and crop damage combined).
Let us begin with the property damage.
By looking at the PROPDMGEXP exponents;
```{r}
unique(data$PROPDMGEXP)
```

We see some lower case characters for example "m" instead of "M". We also see some symbols such as "+". We shall change lower case characters to upper case characters and also set  all symbols to zero character (since the class of the exponents is character)
```{r}
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$PROPDMGEXP[data$PROPDMGEXP %in% c("", "+", "?", "-")] <-  "0"
```

Since PROPDMGEXP stands for the power of 10, we need to convert "B" (for billions) to 9, "M"( for millions) to 6, "K" ( for thousands) to 3  and "H" (for hundreds) to 2.
```{r}
data$PROPDMGEXP[data$PROPDMGEXP == "B"] <- "9"
data$PROPDMGEXP[data$PROPDMGEXP == "M"] <- "6"
data$PROPDMGEXP[data$PROPDMGEXP == "K"] <- "3"
data$PROPDMGEXP[data$PROPDMGEXP == "H"] <- "2"
```

We shall now convert the character class of PROPDMGEXP variable to numeric class and calculating  the true damage by multiplying the damage by the corresponding exponent.Then we shall arrange true.property.damage variable in descending order to find out which even shows the greatest damage in terms of economic loss.
```{r results='hide'}
library(dplyr)
data$PROPDMGEXP <- 10^(as.numeric(data$PROPDMGEXP))
damage <- mutate(data, true.property.damage = PROPDMG*PROPDMGEXP) 
damage1 <- aggregate(damage$true.property.damage, by = list(damage$EVTYPE), FUN = sum)%>%  rename(Event.type = "Group.1", total.property.damage = "x") %>% 
  arrange(desc(total.property.damage))
```


Likewise, let us find out which event type shows the biggest economic loss in case of crops.
Let us just continue with the damage1 dataset
```{r}
unique(damage$CROPDMGEXP)
```
Converting lower case characters to uppercase characters and also setting symbols to zero
```{r}
damage$CROPDMGEXP <- toupper(damage$CROPDMGEXP)
damage$CROPDMGEXP[damage$CROPDMGEXP %in% c("", "+", "?", "-")] <-  "0"

```

Since CROPDMGEXP stands for the power of 10, we need to convert ‘B’ (for billions) to 9, ‘M’( for millions) to 6, ‘K’ ( for thousands) to 3 and ‘H’(for hundreds) to 2;
```{r}
damage$CROPDMGEXP[damage$CROPDMGEXP == "B"] <- "9"
damage$CROPDMGEXP[damage$CROPDMGEXP == "M"] <- "6"
damage$CROPDMGEXP[damage$CROPDMGEXP == "K"] <- "3"
damage$CROPDMGEXP[damage$CROPDMGEXP == "H"] <- "2"

```


converting the character class of CROPDMGEXP variable to numeric class and calculating the true damage by multiplying the damage by the corresponding exponent.
```{r results='hide'}
library(dplyr)
damage$CROPDMGEXP <- 10^(as.numeric(damage$CROPDMGEXP))
damagex <- mutate(damage, true.crop.damage = CROPDMG*CROPDMGEXP)
damage2 <- aggregate(damagex$true.crop.damage, by = list(damagex$EVTYPE), FUN = sum) %>% 
  rename(Event.type = "Group.1", total.crop.damage = "x") %>% 
  arrange(desc(total.crop.damage))
```


Finding the total damage by getting the sum of damage to property and that of crops
```{r results='hide'}
Total_damage <- merge(damage1,damage2) %>% 
  mutate(total.damage = total.property.damage + total.crop.damage) %>% 
  arrange(desc(total.damage)) %>%
  select(c("Event.type", "total.damage"))
#plot
library(ggplot2)
bar_plot_total_damage <- ggplot(Total_damage[1:5,], aes(Event.type, total.damage))+
  geom_bar(aes(fill = Event.type), stat = "identity") + 
  labs(x = "Type of event", 
       y = "Total Damage", 
       title = "Total damage by Event type",
       caption = "Figure 3 : Only the first 5 devastating event types are shown.
       Flood shows the highest damage followed by Hurricane/Typhoon, Tornado, Storm surge,
       and lastly Hail." )+
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(size = 8.5, hjust = 0.5))

```






## **Results**

### *Question 1*

For the event type causing most injuries;
```{r}
head(injuries)
```

we can see that TORNADO causes the most injuries (over 91,000).So we can conclude that if we consider injuries, TORNADO is the most harmful event with respect to population health. This is further shown in the figure below (only the first 5 events are shown).

```{r}
#plot
print(bar_plot_injuries)
```

For the event type causing most fatalities;
```{r}
head(fatalities)
```
we can see that still TORNADO causes the most fatalities(over 5 thousand dollars). So, we can conclude that if we consider fatalities, TORNADO is still the most harmful event with respect to population health.The figure below shows this more clearly (only 5 event types are shown)
```{r}
# plot
print(bar_plot_fatalities)
```
**Conclusion** : After considering the injuries and fatalities caused by the event types, it is realized that TORNADO is the most harmful one with respect to population health across  United States.


### *Question 2*
```{r}
head(damage1)
data.frame(Event.type = damage1$Event.type[which.max(damage1$total.property.damage)],
Amount.dollars = damage1$total.property.damage[which.max(damage1$total.property.damage)])

```
we see that Flood shows the greatest economic loss when it comes to property damage with over 144 billion dollars!

```{r}
head(damage2)
data.frame(Event.type= damage2$Event.type[which.max(damage2$total.crop.damage)],
           Amount.dollars = damage2$total.crop.damage[which.max(damage2$total.crop.damage)])
```
we can see that Drought shows the greatest economic loss when it comes to crop damage with over 13 billion dollars!
```{r}
head(Total_damage)
data.frame(Event.type= Total_damage$Event.type[which.max(Total_damage$total.damage)],
           Amount.dollars = Total_damage$total.damage[which.max(Total_damage$total.damage)])
```
We can see that Flood causes the most damage with a huge economic loss of over 150 billion dollars. This can be shown more clearly by looking at the figure below:
```{r}
print(bar_plot_total_damage)
```

**Conclusion** : After combining the economic loss due to property and crop damage, Flood event emerges as the most devastating event with greatest economic consequences across the United States!*


